Discrete Ordinates

### 2017-07-05 docprep/chapmeth.tex
Added a paragraph about flux moments
Fixed typos and updated LaTeX template

\subsection{The Boltzmann Equation}

@@

\subsection{Angular Discretization}
The coordinate system used in DOCTORS is shown in Fig.~\ref{fig:coord_sys}. A discrete set of $N_a$ angles ($\Omega_{a}, \quad a = 0 \ldots N_a-1$) is selected to represent continuous directional space. Particles are transported only along these discrete directions from each voxel to adjacent voxels.

@@

The $S_N$ quadrature ensures 90 degree symmetry about any axis. In order to ensure the rotational symmetry, the $S_N$ quadrature has a single degree of freedom. The $LQ_N$ quadrature (which is the specific $S_N$ implemented in DOCTORS) uses the remaining degree of freedom to minimize error caused when integrating Legendre polynomials since that class of polynomials are used to represent anisotropy in scattering (see Sec.~\ref{sec:aniso}). For a more through discussion of quadrature generation, see \citet{ref:lewise}. Figure~\ref{fig:quad} shows the simples three quadratures implemented in DOCTORS.

@@

\subsection{Energy Discretization}

@@

The user must provide cross section data in the form of an AMPX formatted binary file (Format detaisl are provided in Section~\ref{sec:xsparse}). The cross section data file contains group structure information. Group averaged cross section values from energy $E_1$ to $E_2$ are computed using Eq.~\ref{eq:groupxs} where $\sigma(E)$ is the continuous microscopic cross section. A weighting function, $f$, is used to weight the energies. The weighting function can be any of a number commonly used functions. The SCALE data libraries use a uniform function [CITE].
\begin{equation}\label{eq:groupxs}
\sigma_G = \frac{\int_{E_1}^{E_2}f(E')\sigma(E') dE'}{\int_{E_1}^{E_2} f(E') dE'}
\end{equation}

In photon based problems, upscatter is typically negligible [CITE]. By the time photons become low enough in energy to upscatter significantly, they are no longer of dosimetric concern. Removing upscatter reduces Eq.~\ref{eq:boltz_e} to
\begin{equation} \label{eq:boltz_e2}
\left[ \hat{\Omega}_a \cdot \nabla + \Sigma_t^g(\boldsymbol{r}) \right]
\psi_{a}^{g}(\boldsymbol{r}) = 
\sum_{a=0}^{N_a-1} \sum_{g'=g}^{G-1} \Sigma_{s, a, a'}^{g, g'}(\boldsymbol{r}) \psi_{a'}^{g'}(\boldsymbol{r}) \omega_a + S_a^g(\boldsymbol{r})
\end{equation}
which nearly identical; only the starting point for the scatter summation changes. Computationally, though, the lack of upscatter makes a significant difference. The highest energy group can be solved independently of all those below it. Once that group is solved, the next group can also be solved and so forth. In the presence of upscatter, the highest group depends on the solution to all groups below it so all energy groups must be solved simultaneously.

@@

\subsection{Spatial Discretization}

The problem domain is split into an evenly spaced Cartesian grid. Uniform grid spacing is not required by discrete ordinate methods though it is helpful and can speed up the computation[CITE].

@@

Voxels can be indexed in one of two ways, either by their component indices or global index. Each voxel in the mesh has three components indices, one in each major direction, $i_x$, $i_y$, and $i_z$ that range from 0 to $N_x - 1$, $N_y - 1$, or $N_z-1$. However, this makes indexing cumbersome, so they are combined into a unique global index, $i$ using 
\begin{equation} \label{eq:indx_flat}
i = i_x (N_z + N_y) + i_y N_z + i_z.
\end{equation}
In general, this is known as "flattening" a matrix into a one-dimensional vector. More details about the flattening implementation used in DOCTORS is given in Section~\ref{sec:flatten}. Using global indexing discussed, the fully discretized the steady state LBE given is
\begin{equation} \label{eq:boltz_i}
\left[ \hat{\Omega}_a \cdot \nabla + \Sigma_{t,i}^g \right]
\psi_{i,a}^{g} = 
\sum_{a=0}^{N_a-1} \sum_{g'=g}^{G-1} \Sigma_{s, i, a, a'}^{g, g'} \psi_{i, a'}^{g'} \omega_a + S_{i,a}^g.
\end{equation}

@@

\subsection{Diamond Difference Approximation}

@@

\subsection{Convergence Criteria}
The typical convergence criteria used to determine whether or not to terminate further iteration is the maximum relative error from the previous iteration to the current iteration given in Eq.~\ref{eq:conv} over the phase space $\mathbb{R}$ where $t$ is the current iteration number. Once the relative error in the parameter of interest, $\phi$, is below some threshold, $\epsilon$, the iteration stops and the solution is considered converged.

@@

ADD FIGURE TO SHOW CONVERGENCE RATE.

@@

\subsection{Anisotropy Treatment}\label{sec:aniso}

INSERT WHY NOT USE FLUX MOMENTS

@@

The data files containing the scatter cross sections stor the information using Legendre polynomial expansion coefficients. The use of a Legendre expansion removes the dependence on the quadrature from the data. This allows any quadrature to work with any cross section dataset. A $l$-order Legendre polynomial is denoted by $P_l$. Details of the Legendre polynomials can be found in Appendix~\ref{appdx:leg}. The anisotropic scatter cross section is rewritten using a Legendre expansion as
\begin{equation} \label{eq:leg_1}
\Sigma_{s, i, a, a'}^{g, g'} = \sum_{l=0}^L \frac{2l+1}{4 \pi}\Sigma_{s, i, l}^{g, g'} P_l(\mu_s).
\end{equation}

An alternate anisotropy treatment is to use the Klein-Nishina formula~\citep{ref:shultisj}. The Klein-Nishina formula describes the scattering cross section as
\begin{equation}
\sigma(E, \mu) \propto q^3 + q\mu^2
\end{equation}
where
\begin{equation}
q(E, \mu) = \frac{1}{1 + (E/E_0)(1-\mu)}
\end{equation}

@@

A rigorous derivation for the uncollided flux is not provided in this work but can be obtained from [CITE]. Instead, an intuitive approach is used. First, consider an isotropic point source in a homogeneous media. The (scalar) flux is well known to be
\begin{equation}
\varphi(r, E) = \frac{S(E) e^{-\mu(E) r}}{4 \pi r^2}
\end{equation}
where $S$ is the source strength, $\mu$ is the attenuation coefficient in the material, and $r$ is the distance traveled by the photon [CITE]. Expanding the problem to a full 3D model, the flux at $\boldsymbol{r}$ from a point source at $\boldsymbol{s}$ is
\begin{equation}\label{eq:isouncol}
\varphi(\boldsymbol{r}, E) = \frac{S(E) e^{-\mu(E) |\boldsymbol{r}-\boldsymbol{s}|}}{4 \pi |\boldsymbol{r}-\boldsymbol{s}|^2}.
\end{equation}
The only non-intuitive step is the introduction of anisotropy. The $4 \pi$ solid angle term in the denominator of Eq.~\ref{eq:isouncol} is generally expressed by
\begin{equation}\label{eq:anisonorm}
\int_{4 \pi}^{} \int_{0}^{\infty} S(E, \hat{\Omega}) dE d\hat{\Omega}
\end{equation}
but DOCTORS assumes $S$ is normalized such that the integral in Eq.~\ref{eq:anisonorm} is $4 \pi$. The flux is nonzero \textit{only} in the direction of travel, namely in the unit direction $\boldsymbol{r} - \boldsymbol{s}$. Therefore, a Kronecker delta function (defined in Eq.~\ref{eq:kronecker}) is used to produce the angular flux:
\begin{equation}\label{eq:uncsol}
\psi(\boldsymbol{r}, E, \hat{\Omega}) = \frac{S(E, \hat{\Omega}) e^{-\mu(E) |\boldsymbol{r}-\boldsymbol{s}|}}{4 \pi |\boldsymbol{r}-\boldsymbol{s}|^2} \delta \left( \hat{\Omega}, \frac{\boldsymbol{r} - \boldsymbol{s}}{|\boldsymbol{r} - \boldsymbol{s}|^2}\right).
\end{equation}

@@

\section{Source Generation}
All photon transport problems considered in the DOCTORS code are ultimately driven by an external source, namely the medical equipment producing the photon beam. These sources can come in a variety of geometric shapes and energy spectra. Currently, DOCTORS has a number of built-in beam types available for analysis including point sources, fan beams, and cone beams. Fan beams and cone beams can be arranged about a phantom to produce an axial slice. Currently, helical paths are not supported, though they could be added with little additional effort. The geometry of fan beams and cone beams in DOCTORS is shown in Fig.~\ref{fig:shape}.

@@

\subsection{Point Source}
The simplest source implemented is an isotropic point source. This source is defined only by its position and energy distribution. Given a vector from the source to the center of the geometry, $\boldsymbol{s}$, and a vector from the source to the center of a cell, $\boldsymbol{c}$,the uncollided flux is computing using Eq.~\ref{eq:isouncol}. The collided flux is then computed using the discrete ordinate solution to the LBE given in Eq.~\ref{eq:col2}. The vectors are shown in Fig.~\ref{fig:phi}.

@@

\subsection{Fan Source}
In addition to the values defining a point source (position $\boldsymbol{s}$ and the energy distribution) a fan source is defined by two additional parameters, $\varphi$ and $\theta$ which describe the azimuthal and polar angles subtended by the fan beam respectively. Figure~\ref{fig:shape_fanbeam} shows $\varphi$ and $\theta$ with respect to the geometry setup. The raytracing algorithm used for the fan beam is identical to the point source raytracer except for one difference: before the raytracing is done, whether the ray falls within the fan beam is determined and any ray falling outside the fan beam is not traced, instead its uncollided flux is immediately set to zero. Some voxels are partially enclosed inside the beam which is not accounted for resulting in artifacts at the edges of the beam.

@@

\subsection{Multi-fan Source}
The focal point about which the CT system rotates is assumed to be the center of the $xy$ plane and at the same $z$-elevation as the first user specified point. The focal spot defines the center of each fan beam.

@@

\subsection{Cone Source and Multi-cone Source}
Cone sources are identical to fan beam sources except for the rejection criteria used to determine whether or not a voxel is inside the beam. The rejection criteria is a function of only a single parameter, $\theta$, the maximum permissible angle between the center of the beam, $\boldsymbol{s}$, and a ray toward the point of interest, $\boldsymbol{c}$. The angle between $\boldsymbol{s}$ and $\boldsymbol{c}$ is
\begin{equation}\label{eq:zetacos}
\cos(\zeta) = \frac{<S_x, S_y, S_z>}{\sqrt{S_x^2 + S_y^2 + S_z^2}} \cdot \frac{<C_x, C_y, C_z>}{\sqrt{C_x^2 + C_y^2 + C_z^2}}
\end{equation}
which, if $\boldsymbol{s}$ and $\boldsymbol{c}$ are unit vectors, simplifies to
\begin{equation}\label{eq:zetacos}
\cos(\zeta) = S_x C_x + S_y C_y + S_z C_z.
\end{equation}
Once the angle is computed, the condition
\begin{equation}\label{eq:zetacon}
|\zeta| < \theta
\end{equation}
is evaluated to determine whether the ray is accepted or rejected. Figure~\ref{fig:beamfancone} shows a comparison of the cone beam and fan beam shapes. The most noticable difference is in the $xz$ plne where the cone beam is circular, but the fan beam produces a rectangular cross section.

@@

\subsection{Fluence-to-dose Conversion}

@@

\subsection{Energy Deposition}

>>>>

\subsection{The Boltzmann equation}\label{sec:lbe}

@@

\subsection{Angular discretization}
The coordinate system used in DOCTORS is shown in Fig.~\ref{fig:coord_sys}. A discrete set of $N_a$ angles ($\Omega_{a}, \quad a = 0 \ldots N_a-1$) is selected to represent continuous directional space. Particles are transported only along these discrete directions from each voxel to adjacent voxels.

@@

The $S_N$ quadrature ensures 90 degree symmetry about any axis. In order to ensure the rotational symmetry, the $S_N$ quadrature has a single degree of freedom. The $LQ_N$ quadrature (which is the specific $S_N$ implemented in DOCTORS) uses the remaining degree of freedom to minimize error caused when integrating Legendre polynomials since that class of polynomials are used to represent anisotropy in scattering (see Sec.~\ref{sec:aniso}). For a more thorough discussion of quadrature generation, see \citet{ref:lewise}. Figure~\ref{fig:quad} shows the simplest three quadratures implemented in DOCTORS.

@@

\subsection{Energy discretization}

@@

The user must provide cross section data in the form of an AMPX formatted binary file (Format details are provided in Section~\ref{sec:xsparse}). The cross section data file contains group structure information. Group averaged cross section values from energy $E_1$ to $E_2$ are computed using Eq.~\ref{eq:groupxs} where $\sigma(E)$ is the continuous microscopic cross section. A weighting function, $f$, is used to weight the energies. The weighting function can be any of a number commonly used functions. The SCALE data libraries use a uniform function.
\begin{equation}\label{eq:groupxs}
\sigma_G = \frac{\int_{E_1}^{E_2}f(E')\sigma(E') dE'}{\int_{E_1}^{E_2} f(E') dE'}
\end{equation}

In photon based problems, upscatter is typically negligible. By the time photons become low enough in energy to upscatter significantly, they are no longer of dosimetric concern. Removing upscatter reduces Eq.~\ref{eq:boltz_e} to
\begin{equation} \label{eq:boltz_e2}
\left[ \hat{\Omega}_a \cdot \nabla + \Sigma_t^g(\boldsymbol{r}) \right]
\psi_{a}^{g}(\boldsymbol{r}) = 
\sum_{a=0}^{N_a-1} \sum_{g'=g}^{G-1} \Sigma_{s, a, a'}^{g, g'}(\boldsymbol{r}) \psi_{a'}^{g'}(\boldsymbol{r}) \omega_a + S_a^g(\boldsymbol{r})
\end{equation}
which nearly identical; only the starting point for the scatter summation changes. Computationally, though, the lack of upscatter makes a significant difference. The highest energy group can be solved independently of all those below it. Once that group is solved, the next group can also be solved and so forth. In the presence of upscatter, the highest group depends on the solution to all groups below it so all energy groups must be solved simultaneously.

@@

\subsection{Spatial discretization}

The problem domain is split into an evenly spaced Cartesian grid. Uniform grid spacing is not required by discrete ordinate methods though it is helpful and can speed up the computation.

@@

Voxels can be indexed in one of two ways, either by their component indices or global index. Each voxel in the mesh has three components indices, one in each major direction, $i_x$, $i_y$, and $i_z$ that range from 0 to $N_x - 1$, $N_y - 1$, or $N_z-1$. However, this makes indexing cumbersome, so they are combined into a unique global index, $i$ using 
\begin{equation} \label{eq:indx_flat}
i = i_x (N_z + N_y) + i_y N_z + i_z.
\end{equation}
In general, this is known as "flattening" a matrix into a one-dimensional vector. More details about the flattening implementation used in DOCTORS is given in Section~\ref{sec:flatten}. Using global indexing discussed, the fully discretized steady state LBE is
\begin{equation} \label{eq:boltz_i}
\left[ \hat{\Omega}_a \cdot \nabla + \Sigma_{t,i}^g \right]
\psi_{i,a}^{g} = 
\sum_{a=0}^{N_a-1} \sum_{g'=g}^{G-1} \Sigma_{s, i, a, a'}^{g, g'} \psi_{i, a'}^{g'} \omega_a + S_{i,a}^g.
\end{equation}

@@

\subsection{Diamond difference approximation}

@@

\subsection{Convergence criteria}
The typical convergence criteria used to determine whether or not to terminate further iteration is the maximum relative error from the previous iteration to the current iteration given in Eq.~\ref{eq:conv} over the phase space $\mathbb{R}$ where $t$ is the current iteration number. Once the relative error in the parameter of interest, $\phi$, is below some threshold, $\epsilon$, the iteration stops and the solution is considered converged.

@@

%ADD FIGURE TO SHOW CONVERGENCE RATE.

@@

\subsection{Anisotropy treatment}\label{sec:aniso}

@@

Typically, at this point, discrete ordinate methods will describe the discrete angular flux, $\psi$ using flux moments, $\phi$. The motivation for using flux moments is that they improve computational performance and reduce memory overhead.

The data files containing the scatter cross sections store the information using Legendre polynomial expansion coefficients. The use of a Legendre expansion removes the dependence on the quadrature from the data. This allows any quadrature to work with any cross section dataset. A $l$-order Legendre polynomial is denoted by $P_l$. Details of the Legendre polynomials can be found in Appendix~\ref{appdx:leg}. The anisotropic scatter cross section is rewritten using a Legendre expansion as
\begin{equation} \label{eq:leg_1}
\Sigma_{s, i, a, a'}^{g, g'} = \sum_{l=0}^L \frac{2l+1}{4 \pi}\Sigma_{s, i, l}^{g, g'} P_l(\mu_s).
\end{equation}

An alternate anisotropy treatment is to use the Klein-Nishina formula~\citep{ref:shultisj}. The Klein-Nishina formula describes the scattering cross section as
\begin{equation}\label{eq:klein}
\sigma(E, \mu) \propto q^3 + q\mu^2
\end{equation}
where
\begin{equation}
q(E, \mu) = \frac{1}{1 + (E/E_0)(1-\mu)}.
\end{equation}
Figure~\ref{fig:klein} shows a comparison of the Klein-Nishina scattering cross section as a function of scatter angle versus the computed cross sections in DOCTORS using the Klein-Nishina formula given in Eq.~\ref{eq:klein}. The slight discrepancy between the two is due to the postprocessing used to merge data from the DOCTORS output. The underlying trends are identical.

\begin{figure}[tb]
  \begin{center}
   \includegraphics[width=3.75in]{figs/klein}
  \end{center}
  \caption{Comparison of the Klein-Nishina formula to the computed cross section values in DOCTORS using the Klein-Nishina anisotropy treatment.}
\label{fig:klein}
\end{figure}

@@

A rigorous derivation for the uncollided flux is not provided in this work. Instead, an intuitive approach is used. First, consider an isotropic point source in a homogeneous media. The (scalar) flux is well known to be
\begin{equation}
\varphi(r, E) = \frac{S(E) e^{-\mu(E) r}}{4 \pi r^2}
\end{equation}
where $S$ is the source strength, $\mu$ is the attenuation coefficient in the material, and $r$ is the distance traveled by the photon [CITE]. Expanding the problem to a full 3D model, the flux at $\boldsymbol{r}$ from a point source at $\boldsymbol{s}$ is
\begin{equation}\label{eq:isouncol}
\varphi(\boldsymbol{r}, E) = \frac{S(E) e^{-\mu(E) |\boldsymbol{r}-\boldsymbol{s}|}}{4 \pi |\boldsymbol{r}-\boldsymbol{s}|^2}.
\end{equation}
The only non-intuitive step is the introduction of anisotropy. The $4 \pi$ solid angle term in the denominator of Eq.~\ref{eq:isouncol} is generally expressed by
\begin{equation}\label{eq:anisonorm}
\int_{4 \pi}^{} \int_{0}^{\infty} S(E, \hat{\Omega}) dE d\hat{\Omega}
\end{equation}
but DOCTORS assumes $S$ is normalized such that the integral in Eq.~\ref{eq:anisonorm} is $4 \pi$. The flux is nonzero \textit{only} in the direction of travel, namely in the unit direction $\boldsymbol{r} - \boldsymbol{s}$. Therefore, a Kronecker delta function (defined in Eq.~\ref{eq:kronecker}) is used to produce the angular flux:
\begin{equation}\label{eq:uncsol}
\psi(\boldsymbol{r}, E, \hat{\Omega}) = \frac{S(E, \hat{\Omega}) e^{-\mu(E) |\boldsymbol{r}-\boldsymbol{s}|}}{4 \pi |\boldsymbol{r}-\boldsymbol{s}|^2} \delta \left( \hat{\Omega}, \frac{\boldsymbol{r} - \boldsymbol{s}}{|\boldsymbol{r} - \boldsymbol{s}|^2}\right).
\end{equation}

@@

\section{Source Generation}\label{sec:source_gen}
All photon transport problems considered in the DOCTORS code are ultimately driven by an external source, namely the medical equipment producing the photon beam. These sources can come in a variety of geometric shapes and energy spectra. Currently, DOCTORS has a number of built-in beam types available for analysis including point sources, fan beams, and cone beams. Fan beams and cone beams can be arranged about a phantom to produce an axial slice. Currently, helical paths are not supported, though they could be added with little additional effort. The geometry of fan beams and cone beams in DOCTORS is shown in Fig.~\ref{fig:shape}.

@@

\subsection{Point source}
The simplest source implemented is an isotropic point source. This source is defined only by its position and energy distribution. Given a vector from the source to the center of the geometry, $\boldsymbol{s}$, and a vector from the source to the center of a cell, $\boldsymbol{c}$,the uncollided flux is computing using Eq.~\ref{eq:isouncol}. The collided flux is then computed using the discrete ordinate solution to the LBE given in Eq.~\ref{eq:col2}. The vectors are shown in Fig.~\ref{fig:phi}.

@@

\subsection{Fan source}
In addition to the values defining a point source (position $\boldsymbol{s}$ and the energy distribution) a fan source is defined by two additional parameters, $\varphi$ and $\theta$ which describe the azimuthal and polar angles subtended by the fan beam respectively. Figure~\ref{fig:shape_fanbeam} shows $\varphi$ and $\theta$ with respect to the geometry setup. The raytracing algorithm used for the fan beam is identical to the point source raytracer except for one difference: before the raytracing is done, whether the ray falls within the fan beam is determined and any ray falling outside the fan beam is not traced, instead its uncollided flux is immediately set to zero. Some voxels are partially enclosed inside the beam which is not accounted for resulting in artifacts at the edges of the beam.

@@

\subsection{Multi-fan source}
The focal point about which the CT system rotates is assumed to be the center of the $xy$ plane and at the same $z$-elevation as the first user specified point. The focal spot defines the center of each fan beam.

@@

\subsection{Cone source and multi-cone source}
Cone sources are identical to fan beam sources except for the rejection criteria used to determine whether or not a voxel is inside the beam. The rejection criteria is a function of only a single parameter, $\theta$, the maximum permissible angle between the center of the beam, $\boldsymbol{s}$, and a ray toward the point of interest, $\boldsymbol{c}$. The angle between $\boldsymbol{s}$ and $\boldsymbol{c}$ is
\begin{equation}\label{eq:zetacos}
\cos(\zeta) = \frac{<S_x, S_y, S_z>}{\sqrt{S_x^2 + S_y^2 + S_z^2}} \cdot \frac{<C_x, C_y, C_z>}{\sqrt{C_x^2 + C_y^2 + C_z^2}}
\end{equation}
which, if $\boldsymbol{s}$ and $\boldsymbol{c}$ are unit vectors, simplifies to
\begin{equation}\label{eq:zetacos}
\cos(\zeta) = S_x C_x + S_y C_y + S_z C_z.
\end{equation}
Once the angle is computed, the condition
\begin{equation}\label{eq:zetacon}
|\zeta| < \theta
\end{equation}
is evaluated to determine whether the ray is accepted or rejected. Figure~\ref{fig:beamfancone} shows a comparison of the cone beam and fan beam shapes. The most noticable difference is in the $xz$ plne where the cone beam is circular, but the fan beam produces a rectangular cross section.

@@

\subsection{Fluence-to-dose conversion}

@@

\subsection{Energy deposition}
